AFQ-Insight

### 2019-12-16 reports/manuscript/abstract.tex

The white matter contains long-range connections between different
brain regions and the organization of these connections hold important
implications for brain function in health and disease. Tractometry
uses diffusion-weighted magnetic resonance imaging (dMRI) data
to quantify tissue properties (e.g. fractional anisotropy (FA),
mean diffusivity (MD), etc.), along the trajectories of these
connections\cite{yeatman2012tract}. Statistical inference from
tractometry data usually either (a) averages these quantities along
the length of the tract in each individual, or (b) performs analysis
point-by-point along each tract, with group comparisons or regression
models computed separately for each point along every one of the tracts.
In the present work, we developed a method based on the sparse group
lasso (SGL) \cite{simon2013sparse} that takes into account tissue
properties measured along all of the tracts, and selects informative
features by enforcing sparsity, not only at the level of individual
tracts, but also across the entire set of tracts and all of the measured
tissue properties. The sparsity penalties for each of these constraints
is identified using a nested cross-validation scheme that guards
against over-fitting and simultaneously identifies the correct
level of sparsity. We demonstrate the accuracy of the method in two
settings: i) In a classification setting, patients with amyotrophic
lateral sclerosis (ALS) are accurately distinguished from matched
controls \cite{sarica2017corticospinal}. Furthermore, SGL automatically
identifies FA in the corticospinal tract as important for this
classification -- correctly finding the parts of the white matter known
to be affected by the disease. ii) In a regression setting, dMRI is
used to accurately predict ``brain age'' \cite{yeatman2014lifespan,
Brown2012-so}. In this case, the weights are distributed throughout the
white matter indicating that many different regions of the white matter
change with development and contribute to the prediction of age. Thus,
SGL makes it possible to leverage the multivariate relationship between
diffusion properties measured along multiple tracts to make accurate
predictions of subject characteristics while simultaneously discovering
the most relevant features of the white matter for the characteristic of
interest.

>>>>

The white matter contains long-range connections between different
brain regions and the organization of these connections hold important
implications for brain function in health and disease. Tractometry
uses diffusion-weighted magnetic resonance imaging (dMRI) data
to quantify tissue properties (e.g. fractional anisotropy (FA),
mean diffusivity (MD), etc.), along the trajectories of these
connections\cite{yeatman2012tract}. Statistical inference from
tractometry data usually either (a) averages these quantities along
the length of the bundle in each individual, or (b) performs analysis
point-by-point along each bundle, with group comparisons or regression
models computed separately for each point along every one of the bundles.
In the present work, we developed a method based on the sparse group
lasso (SGL) \cite{simon2013sparse} that takes into account tissue
properties measured along all of the bundles, and selects informative
features by enforcing sparsity, not only at the level of individual
bundles, but also across the entire set of bundles and all of the measured
tissue properties. The sparsity penalties for each of these constraints
is identified using a nested cross-validation scheme that guards
against over-fitting and simultaneously identifies the correct
level of sparsity. We demonstrate the accuracy of the method in two
settings: i) In a classification setting, patients with amyotrophic
lateral sclerosis (ALS) are accurately distinguished from matched
controls \cite{sarica2017corticospinal}. Furthermore, SGL automatically
identifies FA in the corticospinal tract as important for this
classification -- correctly finding the parts of the white matter known
to be affected by the disease. ii) In a regression setting, dMRI is
used to accurately predict ``brain age'' \cite{yeatman2014lifespan,
Brown2012-so}. In this case, the weights are distributed throughout the
white matter indicating that many different regions of the white matter
change with development and contribute to the prediction of age. Thus,
SGL makes it possible to leverage the multivariate relationship between
diffusion properties measured along multiple bundles to make accurate
predictions of subject characteristics while simultaneously discovering
the most relevant features of the white matter for the characteristic of
interest.

### 2019-12-16 reports/manuscript/sections/conclusion.tex

Taken together, these results demonstrate the promise of the group-regularized
regression approach. Even at the scale of dozens of subjects, the results
provided by SGL are both accurate, as well as interpretable
\cite{Murdoch2019-ax}: tractometry capitalizes on domain knowledge to engineer
meaningful features; SGL scores these features based on their relative
importance; enables a visualization of these feature importance scores in the
anatomical coordinate frame of the tracts (e.g., Figures \ref{fig:class-results}
and fig:regress-beta) and provides a means to understand model errors (e.g.,
Figure \ref{fig:class-errors}) . Thus, this multivariate analysis approach both
(a) achieves high cross-validated accuracy for precision medicine applications
of dMRI data and (b) identifies relevant features of brain anatomy that can
further our neuroscientific understanding of clinical disorders.

@@

The results we present here also motivate extensions of the method using more
sophisticated cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the tracts. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles \cite{Rao2014-xm}. For example, a measurement could come from the
corpus callosum, but also from the right hemipshere. This would also require
extending the cost function used here to incorporate these constraints.
Similarly, unsupervised dimensionality reduction of tractometry data (e.g.,
\cite{Chamberland2019-mu}) could also benefit from constraints based on
grouping.

>>>>

Taken together, these results demonstrate the promise of the group-regularized
regression approach. Even at the scale of dozens of subjects, the results
provided by SGL are both accurate, as well as interpretable
\cite{Murdoch2019-ax}: tractometry capitalizes on domain knowledge to engineer
meaningful features; SGL scores these features based on their relative
importance; enables a visualization of these feature importance scores in the
anatomical coordinate frame of the bundles (e.g., Figures \ref{fig:class-results}
and fig:regress-beta) and provides a means to understand model errors (e.g.,
Figure \ref{fig:class-errors}) . Thus, this multivariate analysis approach both
(a) achieves high cross-validated accuracy for precision medicine applications
of dMRI data and (b) identifies relevant features of brain anatomy that can
further our neuroscientific understanding of clinical disorders.

@@

The results we present here also motivate extensions of the method using more
sophisticated cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the bundles. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles \cite{Rao2014-xm}. For example, a measurement could come from the
corpus callosum, but also from the right hemipshere. This would also require
extending the cost function used here to incorporate these constraints.
Similarly, unsupervised dimensionality reduction of tractometry data (e.g.,
\cite{Chamberland2019-mu}) could also benefit from constraints based on
grouping.

### 2019-12-16 reports/manuscript/sections/introduction.tex

\item Mass univariate approaches: In this approach comparisons between
groups or across individuals are done independently at each node
of each tract, for each one of the diffusion metrics available at
that point. This approach is exhaustive, but statistical power is
compromised by a multiple comparison problem. Different approaches
can be taken to resolving this challenge. For example, Colby and
colleagues \cite{colby2012} used a non-parametric resmapling approach to
correct for family-wise error across the different possible comparisons
\cite{Nichols2002-zu, Nichols2003-yy}.

>>>>

\item Mass univariate approaches: In this approach comparisons between
groups or across individuals are done independently at each node
of each bundle, for each one of the diffusion metrics available at
that point. This approach is exhaustive, but statistical power is
compromised by a multiple comparison problem. Different approaches
can be taken to resolving this challenge. For example, Colby and
colleagues \cite{colby2012} used a non-parametric resmapling approach to
correct for family-wise error across the different possible comparisons
\cite{Nichols2002-zu, Nichols2003-yy}.

### 2019-12-16 reports/manuscript/sections/methods.tex
Update reports/manuscript/sections/methods.tex
Co-Authored-By: Ariel Rokem <arokem@gmail.com>
Add more detail to the ALS dataset description.

\item Diffusion MRI from a previous study of the corticospinal
tract (CST) in patients with amyotrophic lateral sclerosis
(ALS \cite{sarica2017corticospinal}), containing data from 24 ALS
patients and 24 demographically matched healthy controls. These data
were measured in a GE Discovery 750 3T MRI scanner at the Institute
of Bioimaging and Molecular Physiology in Catanzaro. Informed consent
was provided as approved by the Ethical Committee of the University
``Magna Graecia'' of Catanzaro. Voxel resolution was \num{2x2x2}
$\text{mm}^3$ and 27 non-colinear directions were measured with a
$b=1000$ $\frac{\text{sec}}{\text{mm}^2}$. Here, both motion correction
and eddy current correction were applied before modeling the diffusion
tensor in each voxel. We will refer to this dataset as ALS.

\item Diffusion MRI data from a previous study of properties of
the white matter across the lifespan \cite{yeatman2014lifespan},
containing dMRI data from 76 subjects with ages 6-50. These data were
measured in a GE Discovery 750 3T MRI scanner at the Stanford Center
for Cognitive and Neurobiological Imaging. The Stanford University
IRB approved the procedures of this study. Informed consent was
obtained from each adult participant, and assent for participation
was provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}$\text{mm}^3$ with 96 non-colinear directions measured with a
$b=2000$ $\frac{\text{sec}}{\text{mm}^2}$ and 30 non-colinear directions
measured with a $b=1000$ $\frac{\text{sec}}{\text{mm}^2}$. Data was
preprocessed to correct for subject motion and the diffusion tensor
model \cite{basser1994mr} was fit in every voxel, using a robust fit
\cite{chang2005restore}. These data were acquired using a dual spin echo
sequence, in which there is sufficient time for eddy currents to subside
between the application of the gradients and the image acquisition, so
no eddy current correction was applied. We will refer to this dataset
as WH.

>>>>

\item Diffusion MRI from a previous study of the corticospinal
tract (CST) in patients with amyotrophic lateral sclerosis
(ALS \cite{sarica2017corticospinal}), containing data from 24 ALS
patients and 24 demographically matched healthy controls. These data
were measured in a GE Discovery 750 3T MRI scanner at the Institute
of Bioimaging and Molecular Physiology in Catanzaro. Informed consent
was provided as approved by the Ethical Committee of the University
``Magna Graecia'' of Catanzaro. Voxel resolution was \num{2x2x2}
$\text{mm}^3$ and 27 non-colinear directions were measured with a
$b=1000$ $\frac{\text{sec}}{\text{mm}^2}$. Data was preprocessed to 
correct for subject motion and for eddy currents. The diffusion tensor
model \cite{basser1994mr} was fit in every voxel.
We will refer to this dataset as ALS.

\item Diffusion MRI data from a previous study of properties of
the white matter across the lifespan \cite{yeatman2014lifespan},
containing dMRI data from 76 subjects with ages 6-50. These data were
measured in a GE Discovery 750 3T MRI scanner at the Stanford Center
for Cognitive and Neurobiological Imaging. The Stanford University
IRB approved the procedures of this study. Informed consent was
obtained from each adult participant, and assent for participation
was provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}$\text{mm}^3$ with 96 non-colinear directions measured with a
$b=2000$ $\frac{\text{sec}}{\text{mm}^2}$ and 30 non-colinear directions
measured with a $b=1000$ $\frac{\text{sec}}{\text{mm}^2}$. These data
were acquired using a dual spin echo sequence, in which there is
sufficient time for eddy currents to subside between the application of
the gradients and the image acquisition, so no eddy current correction
was applied, but motion correction was applied before fitting the
diffusion tensor model \cite{basser1994mr} in every voxel using a robust
fit \cite{chang2005restore}. We will refer to this dataset as WH.
